A general Nullstellensatz for generalized spaces

### 2019-05-04 paper-qcoh.tex
to do

\begin{defn}The \emph{syntactic site}~$\C_\TT$ of a geometric theory~$\TT$ has:
\begin{enumerate}
\item as objects ``geometric formulas in contexts''~$\{x_1\?X_1,\ldots,x_n\?X_n\_
\varphi\}$ where~$\varphi$ is a geometric formula over the signature of~$\TT$
in the displayed context; \smallskip
\item as set of morphisms~$\Hom_{\C_\TT}(\{\vec x\_
\varphi\}, \{\vec y\_ \psi\})$ the set of formulas~$\theta$ in the
context~$\vec x, \vec y$ which are~$\TT$-provably functional,
modulo~$\TT$-provable equivalence of such formulas; \smallskip
\item as covering families those families~$(\{\vec x_i\_ \varphi_i\} \xrightarrow{\theta_i}
\{\vec y\_ \psi\})_i$ for which~$\TT$ proves the sequent~$(\psi \seq{\vec y} \bigvee_i \exists \vec x_i\_
\theta_i)$.
\end{enumerate}
\end{defn}

@@

Writing~$\yon : \C_\TT \to \Set[\TT]$ for the Yoneda embedding, the \emph{generic
model}~$U_\TT$ of~$\TT$ interprets a sort~$X$ of~$\TT$ as the sheaf~$\yon\{x\?X\_ \top\}$,
a function symbol~$f : X_1 \cdots X_n \to Y$ as the morphism given by
the~$\TT$-provably functional formula~$f(x_1,\ldots,x_n) = y$ and a relation
symbol~$R \rightarrowtail X_1 \cdots X_n$ by the subobject~$\yon\{\vec x\_ R(\vec
x)\} \rightarrowtail \yon\{\vec x\_ \top\}$.

@@

@@

Given an internal signature~$\Sigma$ internal to a Grothendieck topos~$\E$ (or
elementary topos with a natural numbers object), we can successively build the
object of contexts (the object of lists of sorts); the object of terms
(equipped with a morphism to the object of contexts); the object of atomic
propositions (again equipped with such a morphism); the object of geometric
formulas (again so); the object of geometric sequents (again so). A internal
geometric theory~$\TT$ over~$\Sigma$ is then given by a subobject of the object
of geometric sequents, interpreted as the object of axioms of~$\TT$. Given such
an internal theory~$\TT$, we can build the object of proof trees of~$\TT$.

@@

\section{Proofs of the main theorems}
\label{sect:main}

Given a geometric theory~$\TT$ with its generic model~$U_\TT$ in~$\Set[\TT]$,
the main theorems reference a certain internal
theory~$\ul{\TT}/U_\TT$ internal to~$\Set[\TT]$. This theory is defined as
follows.

@@

\begin{rem}From the internal point of view of~$\Set[\TT]$, we can construct the
classifying topos of~$\ul{\TT}/U_\TT$. Externally, this construction gives rise
to a bounded topos over~$\Set[\TT]$, hence to a Grothendieck topos. Using for
instance the technique described
in~\cite{blechschmidt-hutzler-oldenziel:composition}, one can show that this topos classifies the
theory of homomorphisms between~$\TT$-models. It can also be obtained as the
lax pullback~$(\Set[\TT] \Rightarrow_{\Set[\TT]} \Set[\TT])$.\end{rem}

\begin{lemma}\label{lemma:truth-to-provability}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a geometric formula over the
signature of~$\TT$ in a context~$x_1\?X_1,\ldots,x_n\?X_n$. Then
\[ \{ \vec x\_ \alpha \} \models \speak{$\ul{\TT}/U_\TT$ proves $(\top
\vdash_{[]} \alpha)$}, \]
where the free variables~$\vec x$ occurring in~$\alpha$ are interpreted as in
Proposition~\ref{prop:basic-truth}.
\end{lemma}

@@

The resulting formula~$\varphi'$ is a geometric formula over the signature
of~$\TT/U_\TT(A)$.
The verification of~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
$(\varphi \dashv\vdash \varphi')$}$ rests on the observation
\[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_k\_
\theta_k) \seq{[]} \bigvee_\ell \{ \top \,|\, \text{$(\exists \vec y_\ell\_ \theta_\ell)$
holds for~$U_\TT$} \}\bigr)$} \]
which in turn can be checked on the covering~$(\{ \vec y_j\_ \alpha_j \}
\xrightarrow{[\theta_j]} \{ \vec x\_ \alpha \})_j$, applying
Lemma~\ref{lemma:truth-to-provability} and using that~$\TT$ (and
hence~$\ul{\TT}$) proves~$(\exists \vec y_j\_ \theta_j) \wedge (\exists \vec
y_k\_ \theta_k) \seq{\vec x} \bigvee \{ \top \,|\, j = k \}$. (This kind of
reasoning also appears, for instance, in~\cite[p.~20]{blass:seven-trees}.)
\end{proof}

@@

The force of the Nullstellensatz of Theorem~\ref{thm:nullstellensatz} is for
sequents~$\sigma$ which are not of the form~$(\top \seq{[]} \varphi)$. Indeed,
for such sequents, even~$\TT$ proves~$(\speak{$\sigma$ holds for~$U_\TT$}
\Rightarrow \speak{$\ul{\TT}/U_\TT$ proves~$(\top \seq{[]} \varphi)$})$, as can
be seen by a simple induction on the structure of~$\varphi$. For more general
sequents, Theorem~\ref{thm:nullstellensatz} expresses a nontrivial fact: that
truth of a conditional, universally-quantified statement implies a single
unconditional statement.

\begin{rem}Theorem~\ref{thm:nullstellensatz} cannot be strengthened to
arbitrary geometric sequents. For instance, in the case that~$\TT$ is the
theory of objects, the internal geometric sequent~$(\top \seq{x\?U_\TT}
\bigvee_{a\?U_\TT} (x = e_a))$ trivially holds for~$U_\TT$. However, this sequent is not
provable modulo~$\ul{\TT}/U_\TT$, as for instance the model~$U_\TT \amalg
\{\star\}$ does not validate it.\end{rem}

@@

\begin{rem}In the case that the~$\Set[\TT']$ is a dense subtopos
of~$\Set[\TT]$, Theorem~\ref{thm:nullstellensatz-horn} can be strengthened to
allow~$\bot$ as consequent, since in this case~$(\nabla\bot) \Rightarrow \bot$.\end{rem}

@@

Throughout this section, let~$\TT$ be a Horn theory. It is a basic result that
Horn theories are of presheaf type, such that the classifying topos~$\Set[\TT]$
can be taken as the topos of functors~$\Mod{\TT}_\mathrm{fp} \to \Set$,
where~$\Mod{\TT}_\mathrm{fp}$ is the full subcategory of the category
of~$\TT$-models (in~$\Set$) on the finitely presentable objects. The
(underlying object of the) generic model is the tautologous functor~$U_\TT : T
\mapsto T$. By~\emph{$U_\TT$-algebra} we mean, in analogy with the terminology in
the case that~$\TT$ is the theory of rings, a~$\TT$-model~$M$ equipped with
a~$\TT$-homomorphism~$U_\TT \to M$.

@@

\begin{proof}The desired model can be constructed as a term algebra. As a set,
it consists of the terms (in the empty context) of the signature~$\Sigma + X$
modulo the equivalence relation identifying two terms if and only if~$\TT + R$
proves them to be equal. The function symbols~$f$ of~$\Sigma$ are interpreted
by declaring~$\brak{f}([t_1],\ldots,[t_n]) = [f(t_1,\ldots,t_n)]$ and the
relation symbols~$S$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
\brak{S} \Leftrightarrow (\TT + R \vdash S(t_1,\ldots,t_n))$.

We omit the required verifications and only remark that while the same
construction could be carried out if~$\TT$ was a general geometric theory, the
resulting object would in general not be a model of~$\TT$.
\end{proof}

@@

\begin{proof}It is an instructive exercise to verify that models of the stated
form are compact. Conversely, let a~$\TT$-model~$M$ be given. Then~$\TT$ is the
filtered colimit of all models over~$M$ which are of the stated form. If~$M$ is
compact, the identity on~$M$ factors over such a model. Hence~$M$ is a
retract of such a model and hence itself isomorphic to a model of this form.
\end{proof}

@@

\begin{proof}By the Yoneda lemma, the sections of the sheaf~$\Spec(A^\sim) :
\Mod{\TT}_\mathrm{fp} \to \Set$ on an object~$T$ are given by the set
\begin{align*}
  \Spec(A^\sim)(T) &\cong \Hom(\yon T, \Spec(A^\sim)) =
  \Hom(\yon T, [A^\sim,U_\TT]_{U_\TT}) \\
  &\cong \Hom(\yon T \times A^\sim, U_\TT)_{\text{$U_\TT$-algebra homomorphism in second
  argument}} \\
  &\cong \Hom_{U_\TT}(A^\sim, (U_\TT)^{\yon T})
  \cong \Hom_{U_\TT}(A^\sim, U_\TT|T),
\end{align*}
where~$[A^\sim,U_\TT]_{U_\TT}$ is the object of~$U_\TT$-algebra homomorphisms from~$A^\sim$
to~$U_\TT$ (a subobject of the internal Hom~$(U_\TT)^{A^\sim}$); $\Hom_{U_\TT}$
denotes the set of~$U_\TT$-algebra homomorphisms; $(U_\TT)^{\yon T}$ is the object of morphisms
from~$\yon T$ to~$U_\TT$; and~$U_\TT|T$ is the functor~$U_\TT(T \times \cdot)$, that is~$S
\mapsto T \otimes S$.

@@

By \emph{extended geometric logic} we mean the extension of geometric logic
where we are allowed to form, in addition to the basic sorts supplied by
a given signature, finite limits of sorts and set-indexed colimits of sorts. By
(intuitionistic) \emph{higher-order logic}, we mean the further extension where
we may also form powersorts. These derived sorts come with respective term
constructors (tuple formers, coprojections, set comprehension) and the usual
rules governing these constructors.

An \emph{extended geometric formula} is a formula of extended geometric logic built from equality and
relation symbols by the logical
connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$. The existential
quantification can be over any sorts of extended geometric logic, including the derived sorts. An
\emph{extended geometric sequent} is a sequent of the form~$(\varphi
\seq{\vec x} \psi)$ where~$\varphi$ and~$\psi$ are extended geometric
formulas and the sorts of the variables~$\vec x$ may be derived sorts.

@@

\begin{rem}\label{rem:definability-entails-nullstellensatz}
The proof of Theorem~\ref{thm:definability} used the Nullstellensatz
of Theorem~\ref{thm:nullstellensatz}. Conversely, the internal statement of
Theorem~\ref{thm:definability} entails the internal
statement~\eqref{eq:nullstellensatz} of Theorem~\ref{thm:nullstellensatz} in
extended geometric logic. Indeed, leaving the translation to formal logic to
the reader, let~$\sigma = (\varphi \seq{\vec x} \psi)$ be a
geometric$^\star$ sequent over the signature of~$\ul{\TT}/U_\TT$ such
that~$\sigma$ holds for~$U_\TT$. Then the subsets~$\{ (\vec x) \,|\, \varphi
\}$ and~$\{ (\vec x) \,|\, \varphi \wedge \psi \}$ are equal. Hence the
assumption implies that the formulas~$\varphi$
and~$\varphi \wedge \psi$ are provably$^\star$ equivalent.
Thus~$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi \seq{\vec x} \psi)$.
\end{rem}

@@

\begin{thm}\label{thm:higher-order-nullstellensatz}
Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$, for any
extended geometric$^\star$ sequent~$\sigma$ over the signature
of~$\ul{\TT}/U$, the following statements are equivalent:
\begin{enumerate}
\item The sequent~$\sigma$ holds for~$U_\TT$. \smallskip
\item The sequent~$\sigma$ is provable$^\star$ modulo~$\TT/U_\TT$ in extended
geometric logic.
\end{enumerate}
\end{thm}

\begin{proof}The implication~$(2) \Rightarrow (1)$ is
immediate since~$U_\TT$ is a model of~$\TT/U_\TT$. The converse direction is by
Lemma~\ref{lemma:extended-conservative}, which holds internally in~$\Set[\TT]$
as the proof we supplied is constructive, and the Nullstellensatz for ordinary
geometric logic of Theorem~\ref{thm:nullstellensatz}.
\end{proof}

\begin{thm}\label{thm:higher-order-characterization}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a higher-order formula over
the signature of~$\TT$. Then the following statements are equivalent:
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$. \smallskip
\item The formula~$\alpha$ is provable in higher-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axiom scheme
\begin{equation}\label{eq:higher-order-nullstellensatz}\tag{$\P$}
  \speak{the map~$\mathrm{Form}^\star_{\vec
  x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x}) \longrightarrow P(X_1 \times
  \cdots \times X_n)$ is bijective},
\end{equation}
where~$X_1,\ldots,X_n$ is any list of sorts.
\end{enumerate}
\end{thm}

>>>>

\begin{defn}The \emph{syntactic site}~$\C_\TT$ of a geometric theory~$\TT$ has:
\begin{enumerate}
\item as objects \emph{geometric formulas in contexts}~$\{x_1\?X_1,\ldots,x_n\?X_n\_
\varphi\}$ where~$\varphi$ is a geometric formula over the signature of~$\TT$
in the displayed context; \smallskip
\item as set~$\Hom_{\C_\TT}(\{\vec x\_ \varphi\}, \{\vec y\_ \psi\})$ of morphisms
the set of formulas~$\theta$ in the
context~$\vec x, \vec y$ which are~$\TT$-provably functional,
modulo~$\TT$-provable equivalence of such formulas; \smallskip
\item as covering families those families~$(\{\vec x_i\_ \varphi_i\} \xrightarrow{\theta_i}
\{\vec y\_ \psi\})_i$ for which~$\TT$ proves the sequent~$(\psi \seq{\vec y} \bigvee_i \exists \vec x_i\_
\theta_i)$.
\end{enumerate}
\end{defn}

@@

Writing~$\yon : \C_\TT \to \Set[\TT]$ for the Yoneda embedding,\footnote{With
this notational choice we are following Riehl and
Verity~\cite{riehl-verity:elements}.}
the \emph{generic
model}~$U_\TT$ of~$\TT$ interprets a sort~$X$ of~$\TT$ as the sheaf~$\yon\{x\?X\_ \top\}$,
a function symbol~$f : X_1 \cdots X_n \to Y$ as the morphism given by
the~$\TT$-provably functional formula~$f(x_1,\ldots,x_n) = y$ and a relation
symbol~$R \rightarrowtail X_1 \cdots X_n$ by the subobject~$\yon\{\vec x\_ R(\vec
x)\} \rightarrowtail \yon\{\vec x\_ \top\}$.

@@

\begin{proof}See, for instance,~\cite[Theorem~2.1.8]{caramello:tst}
or~\cite[discussion before Proposition~D3.1.12]{johnstone:elephant}.\end{proof}

@@

Given an internal signature~$\Sigma$ internal to a Grothendieck topos~$\E$ (or
elementary topos with a natural numbers object), we can successively build the
object of contexts (the object of lists of sorts); the object of terms
(equipped with a morphism to the object of contexts); the object of atomic
propositions (again equipped with such a morphism); the object of geometric
formulas (again so); and the object of geometric sequents (again so). A internal
geometric theory~$\TT$ over~$\Sigma$ is then given by a subobject of the object
of geometric sequents, interpreted as the object of axioms of~$\TT$. Given such
an internal theory~$\TT$, we can build the object of proof trees of~$\TT$.

@@

\section{The first-order Nullstellensatz}
\label{sect:main}

Given a geometric theory~$\TT$ with its generic model~$U_\TT$ in~$\Set[\TT]$,
our main theorems will reference a certain internal
theory~$\ul{\TT}/U_\TT$ internal to~$\Set[\TT]$. This theory is defined as
follows.

@@

\begin{rem}From the internal point of view of~$\Set[\TT]$, we can construct the
classifying topos of~$\ul{\TT}/U_\TT$. Externally, this construction gives rise
to a bounded topos over~$\Set[\TT]$, hence to a Grothendieck topos. Using for
instance the technique described
in~\cite{blechschmidt-hutzler-oldenziel:composition}, one can show that this topos classifies the
theory of homomorphisms between~$\TT$-models. It can also be obtained as the
lax pullback~$(\Set[\TT] \Rightarrow_{\Set[\TT]} \Set[\TT])$.
There are two canonical geometric morphisms from this topos to~$\Set[\TT]$, the
morphism computing the domain and the morphism computing the codomain; the
morphism obtained by externalizing the internal construction is the former.
\end{rem}

\begin{lemma}\label{lemma:truth-to-provability}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a geometric formula over the
signature of~$\TT$ in a context~$x_1\?X_1,\ldots,x_n\?X_n$. Then
\[ \{ \vec x\_ \alpha \} \models \speak{$\ul{\TT}/U_\TT$ proves $(\top
\vdash_{[]} \alpha)$}, \]
where the free variables~$\vec x$ occurring in~$\alpha$ are first interpreted as in
Proposition~\ref{prop:basic-truth} and then regarded as the induced constant
symbols.
\end{lemma}

@@

The resulting formula~$\varphi'$ is a geometric formula over the signature
of~$\TT/U_\TT(A)$.
The verification of~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
$(\varphi \dashv\vdash \varphi')$}$ rests on the observation
\[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_k\_
\theta_k) \seq{[]} \bigvee_\ell \{ \top \,|\, \text{$(\exists \vec y_\ell\_ \theta_\ell)$
holds for~$U_\TT$} \}\bigr)$} \]
which in turn can be checked on the covering~$(\{ \vec y_j\_ \alpha_j \}
\xrightarrow{[\theta_j]} \{ \vec x\_ \alpha \})_j$, applying
Lemma~\ref{lemma:truth-to-provability} and using that~$\TT$ (and
hence~$\ul{\TT}$) proves~$((\exists \vec y_j\_ \theta_j) \wedge (\exists \vec
y_k\_ \theta_k) \seq{\vec x} \bigvee \{ \top \,|\, j = k \})$. (This kind of
reasoning also appears, for instance, in~\cite[p.~20]{blass:seven-trees}.)
\end{proof}
% XXX check last paragraph; change proves* to proves.

@@

The force of the Nullstellensatz of Theorem~\ref{thm:nullstellensatz} is for
sequents~$\sigma$ which are not of the form~$(\top \seq{[]} \varphi)$. Indeed,
for those special sequents, even~$\TT$ proves~$(\speak{$\sigma$ holds for~$U_\TT$}
\Rightarrow \speak{$\ul{\TT}/U_\TT$ proves~$(\top \seq{[]} \varphi)$})$, as can
be seen by a simple induction on the structure of~$\varphi$. For more general
sequents, Theorem~\ref{thm:nullstellensatz} expresses a nontrivial fact: that
truth of an universally-quantified conditional statement implies a single
unquantified unconditional statement.

\begin{rem}The restriction in Theorem~\ref{thm:nullstellensatz} to
geometric$^\star$ sequents cannot be lifted, that is the generalization
of~Theorem~\ref{thm:nullstellensatz} to
arbitrary internal geometric sequents is false. For instance, in the case that~$\TT$ is the
theory of objects, the internal geometric sequent~$(\top \seq{x\?U_\TT}
\bigvee_{a\?U_\TT} (x = e_a))$ trivially holds for~$U_\TT$. However, this sequent is not
provable modulo~$\ul{\TT}/U_\TT$, as for instance the model~$U_\TT \amalg
\{\star\}$ does not validate it.\end{rem}

@@

\begin{rem}In the case that the~$\Set[\TT']$ is a dense subtopos
of~$\Set[\TT]$, Theorem~\ref{thm:nullstellensatz-horn} can be strengthened to
allow~$\bot$ as consequent, since in this case we have~$(\nabla\psi)
\Rightarrow \psi$ also for~$\psi \equiv \bot$.\end{rem}

@@

Throughout this section, let~$\TT$ be a Horn theory. It is a basic result that
Horn theories are of presheaf type, such that the classifying topos~$\Set[\TT]$
can be taken as the topos of functors~$\Mod{\TT}_\mathrm{fp} \to \Set$,
where~$\Mod{\TT}_\mathrm{fp}$ is the full subcategory of the category
of~$\TT$-models (in~$\Set$) on the finitely presentable objects~\cite[Theorem~2.1.21]{caramello:tst}. The
(underlying object of the) generic model is the tautologous functor~$U_\TT : T
\mapsto T$. By~\emph{$U_\TT$-algebra} we mean, in analogy with the terminology in
the case that~$\TT$ is the theory of rings, a~$\TT$-model~$M$ equipped with
a~$\TT$-homomorphism~$U_\TT \to M$.

% XXX: upgrade to cartesian case? maybe as a remark?

@@

\begin{proof}The desired model can be constructed as a term algebra. As a set,
it consists of the terms (in the empty context) of the signature~$\Sigma + X$
modulo the equivalence relation identifying two terms if and only if~$\TT + R$
proves them to be equal. The function symbols~$f$ of~$\Sigma$ are interpreted
by declaring~$\brak{f}([t_1],\ldots,[t_n]) = [f(t_1,\ldots,t_n)]$ and the
relation symbols~$S$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
\brak{S} \Leftrightarrow (\TT + R \,\vdash S(t_1,\ldots,t_n))$.

We omit the required verifications and only remark that while the same
construction could be carried out if~$\TT$ was a general geometric theory, the
term algebra would in general not be a model of~$\TT$.
\end{proof}

@@

\begin{proof}It is an instructive exercise to verify that models of the stated
form are compact. Conversely, let a~$\TT$-model~$M$ be given. Then~$\TT$ is the
% XXX cite Adamek/Rosicky
filtered colimit of all models over~$M$ which are of the stated form. If~$M$ is
compact, the identity on~$M$ factors over such a model. Hence~$M$ is a
retract of such a model and hence itself isomorphic to a model of this form.
\end{proof}

@@

\begin{proof}By the Yoneda lemma, the sections of the sheaf~$\Spec(A^\sim) :
\Mod{\TT}_\mathrm{fp} \to \Set$ on an object~$T$ are given by the set
\begin{align*}
  \Spec(A^\sim)(T) &\cong \Hom(\yon T, \Spec(A^\sim)) =
  \Hom(\yon T, [A^\sim,U_\TT]_{U_\TT}) \\
  &\cong \Hom(\yon T \times A^\sim, U_\TT)_{\text{$U_\TT$-algebra homomorphism in second
  argument}} \\
  &\cong \Hom_{U_\TT}(A^\sim, (U_\TT)^{\yon T})
  \cong \Hom_{U_\TT}(A^\sim, U_\TT|T),
\end{align*}
where~$[A^\sim,U_\TT]_{U_\TT}$ is the object of~$U_\TT$-algebra homomorphisms from~$A^\sim$
to~$U_\TT$ (a subobject of the internal Hom~$(U_\TT)^{A^\sim}$); $\Hom_{U_\TT}$
denotes the set of~$U_\TT$-algebra homomorphisms; $(U_\TT)^{\yon T}$ is the object of morphisms
from~$\yon T$ to~$U_\TT$; and~$U_\TT|T$ is the functor~$U_\TT(T \times \cdot)$, that is
the functor~$S \mapsto T \otimes S$.

@@

By \emph{extended geometric logic} we mean the extension of geometric logic
where we are allowed to form, in addition to the basic sorts supplied by
a given signature, finite limits of sorts and set-indexed colimits of sorts. By
(intuitionistic) \emph{higher-order logic}, we mean the further extension where
we may also form powersorts. These derived sorts come with respective term
constructors (tuple formers, coprojections, set comprehension) and the usual
rules governing these constructors.
% XXX cite some reference for these "usual rules", perhaps Milly?

An \emph{extended geometric formula} is a formula of extended geometric logic built from equality and
relation symbols by the logical
connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$. Existential
quantification can be over any sorts of extended geometric logic, including the derived sorts. An
\emph{extended geometric sequent} is a sequent of the form~$(\varphi
\seq{\vec x} \psi)$ where~$\varphi$ and~$\psi$ are extended geometric
formulas and the sorts of the variables~$\vec x$ may be derived sorts.

@@

\begin{rem}\label{rem:definability-entails-nullstellensatz}
The proof of Theorem~\ref{thm:definability} used the Nullstellensatz
of Theorem~\ref{thm:nullstellensatz}. Conversely, the internal statement of
Theorem~\ref{thm:definability} entails the internal
statement~\eqref{eq:nullstellensatz} of Theorem~\ref{thm:nullstellensatz} in
% XXX "the internal statement"? be more clear...
extended geometric logic. Indeed, leaving the translation to formal logic to
the reader, let~$\sigma = (\varphi \seq{\vec x} \psi)$ be a
geometric$^\star$ sequent over the signature of~$\ul{\TT}/U_\TT$ such
that~$\sigma$ holds for~$U_\TT$. Then the subsets~$\{ (\vec x) \,|\, \varphi
\}$ and~$\{ (\vec x) \,|\, \varphi \wedge \psi \}$ are equal. Hence the
assumption implies that the formulas~$\varphi$
and~$\varphi \wedge \psi$ are provably$^\star$ equivalent.
Thus~$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi \seq{\vec x} \psi)$.
\end{rem}

@@

\begin{thm}\label{thm:higher-order-nullstellensatz}
Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$, for any
extended geometric$^\star$ sequent~$\sigma$ over the signature
of~$\ul{\TT}/U$, the following statements are equivalent:
\begin{enumerate}
\item The sequent~$\sigma$ holds for~$U_\TT$. \smallskip
\item The sequent~$\sigma$ is provable$^\star$ modulo~$\ul{\TT}/U_\TT$ in extended
geometric logic.
\end{enumerate}
\end{thm}

\begin{proof}The implication~$(2) \Rightarrow (1)$ is
immediate since~$U_\TT$ is a model of~$\ul{\TT}/U_\TT$. The converse direction is by
Lemma~\ref{lemma:extended-conservative}, which holds internally in~$\Set[\TT]$
as the proof we supplied is constructive, and by the Nullstellensatz for ordinary
geometric logic of Theorem~\ref{thm:nullstellensatz}.
\end{proof}

\begin{thm}\label{thm:higher-order-characterization}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a higher-order formula over
the signature of~$\TT$. Then the following statements are equivalent:
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$. \smallskip
\item The formula~$\alpha$ is provable in higher-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axiom scheme
\begin{equation}\label{eq:higher-order-nullstellensatz}\tag{$\P$}
  \speak{the map~$\mathrm{Form}^\star_{\vec
  x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x}) \longrightarrow P(X_1 \times
  \cdots \times X_n)$ is bijective},
\end{equation}
where~$X_1,\ldots,X_n$ is any list of sorts.
% XXX think about variables...
\end{enumerate}
\end{thm}
